The CrypTool Script: \\ Cryptography, Mathematics and more \\ \vspace{30pt} {\large Background reading\\*[-8pt] for the free eLearning program CrypTool (version 1.4.00)} \vspace{60pt}

### 2007-04-23 script/en/chapters/moderncryptography.tex
be changes dated 2006-04-09 2006-04-11 2006-05-05 2006-05-30 2006-07-26

\newpage
%**************************************************************************************************************************
\section{The Mathematical Ideas behind Modern Cryptography}
\label{Chapter_ModernCryptography}
\hypertarget{Chapter_ModernCryptography}{}
%**************************************************************************************************************************
(Oyono R. / Esslinger B., Sep. 2000, Updates Nov. 2000, Feb. 2003)

%--------------------------------------------------------------------------------------------------------------------------
\subsection{One way functions with trapdoor and complexity classes}
%--------------------------------------------------------------------------------------------------------------------------
\index{Cryptography!modern} \index{One way function} \hypertarget{OneWayFunktion1}{}
A {\bf one way function} is a function that can be calculated 
efficiently, but whose inverse is extremely complicated and practically 
impossible to calculate.\par

@@

\newpage
%--------------------------------------------------------------------
\subsection{Knapsack problem as a basis for public key procedures}%
\index{Cryptography!public key}
%--------------------------------------------------------------------
%--------------------------------------------------------------------
\subsubsection{Knapsack problem} \index{Knapsack}
%--------------------------------------------------------------------

@@

\subsection{Decomposition into prime factors as a basis for public key 
procedures}\index{Prime factor!decomposition}

%--------------------------------------------------------------------
\subsubsection[The RSA procedure]
{The RSA procedure\footnotemark}
\footnotetext{
Using CrypTool\index{CrypTool} you can gain practical experience
with the RSA procedure via the menu {\bf Indiv.Procedures \textbackslash{} 
RSA Cryptosystem \textbackslash{} RSA Demonstration}.
}
\index{RSA} \hypertarget{RSAVerfahren}{} 
As early as 1978, R. \index{Rivest, Ronald} Rivest, \index{Shamir, Adi} A. Shamir,
\index{Adleman, Leonard} L. Adleman \cite{RSA1978} introduced 
the most important asymmetric cryptography procedure to date.  \par

@@

If we know the various prime factors $ p_1, \dots , p_k $ of $ N, $ then $ \phi 
(N) = N \cdot (1-\frac{1}{p_1}) \,
\cdots \, (1-\frac{1}{p_k}) $\footnote{Further formulas for the Euler phi
function are in the article ``Introduction to Elementary Number Theory with
Examples'', chapter~\ref{patternsandstructures}.}.

@@

At the moment, no algorithm is known that can factorise two prime numbers 
sufficiently quickly for extremely large values (e.g. for several hundred 
decimal places). The quickest algorithms known today
\cite{4Stinson1995}\index{Stinson 1995} 
factorise a compound whole number $ N $ in a time period proportional to $ L(N)= 
e^{\sqrt{\ln (N) \ln (\ln (N))}}. $ 
\vskip +5 pt
\begin{center}
\begin{tabular}{|l||l|l|l|l|l|l|}\hline
$N$ & $ 10^{50} $ & $ 10^{100} $ & $ 10^{150} $ & $ 10^{200} $ & $ 10^{250} $ & 
$ 10^{300} $ \\ \hline
$L(N)$ & $ 1.42 \cdot 10^{10} $ &  $ 2.34  \cdot 10^{15} $ &  $ 3.26 \cdot 
10^{19} $ &  $ 1.20 \cdot 10^{23} $ &  $ 1.86 \cdot 10^{26} $ &  $ 1.53 \cdot 
10^{29} $ \\ \hline
\end{tabular}
\end{center}
\vskip +5 pt
As long as no better algorithm is found, this means that values of the order of 
magnitude $ 100 $ to $ 200 $ decimal places are currently safe. Estimates of the 
current computer technology indicate that a number with $100$ decimal places 
could be factorised in approximately two weeks at justifiable cost. Using an 
expensive configuration (e.g. of around 10 million US dollars), a number with 
$150$ decimal places could be factorised in about a year. A $200-$digit number 
should remain impossible to factorise for a long time to come, unless there is a 
mathematical breakthrough. For example, it would take about 1000 years to 
decompose a 200-digit number into prime factors using the existing algorithms; 
this applies even if $ 10^{12} $  operations can be performed per second, which 
is beyond the performance\index{Performance} of current technology and would cost billions of 
dollars in development costs. However, you can never be sure that there won't be 
a mathematical breakthrough tomorrow.

To this date, it has not been proved that the problem of breaking RSA is 
equivalent to the factorisation problem. Nevertheless, it is clear that the RSA 
procedure will no longer be safe if the factorisation problem is ``solved''.

\subsubsection{Rabin public key procedure (1979)}

@@

       \subsection{The discrete logarithm as a basis for public key procedures}
Discrete logarithms \index{Discrete logarithm} form the basis for a large number of algorithms for public-
key procedures.
\index{Logarithm problem!discrete}

    \subsubsection{The discrete logarithm in $ \Z_p $}
Let $ p $ be a prime number and let $ g \in \Z_p^\ast=\{0,1,\ldots,p-1\} $. Then 
the discrete exponential function base $ g $  is defined as
\[ e_g : k \longrightarrow y:=g^k \mod p, \quad 1\leq k \leq p-1. \]
The inverse function is called a discrete logarithm function $ \log_g $; the 
following holds:
\[ \log_g (g^k) =k. \]
\index{Exponential function!discrete} The problem of the discrete logarithm (in 
$ \Z_p^\ast$) is understood to be as follows:
\[ {\rm Given~ } p,g {\rm ~and~ } y, {\rm~determine~} k {\rm ~such~that~ } 
y=g^k \mod p {\rm .}\]
It is much more difficult to calculate the discrete logarithm than to evaluate 
the discrete exponential function (see chapter \ref{MultOrdPrimitveRoot}).
There are several procedures for calculating the discrete logarithm 
\cite{4Stinson1995}\index{Stinson 1995}:
\vskip + 5pt
\begin{center}
\begin{tabular}{|l|l|}\hline
Name                 &        Complexity \\ \hline \hline
Baby-Step-Giant-Step &         $ O(\sqrt{p}) $ \\ \hline
Silver-Pohlig-Hellman    &    polynomial in $ q, $ the greatest \\
&  prime factor of $ p-1. $ \\ \hline
Index-Calculus &             $ O(e^{(1+o(1)) \sqrt{\ln (p) \ln (\ln (p))}}) $ \\ 
\hline
\end{tabular}
\end{center}
\vskip +5 pt
\index{Silver} \index{Pohlig, S. C.} \index{Hellman, Martin}
\index{Baby-Step-Giant-Step}

@@

%--------------------------------------------------------------------
\subsubsection{ElGamal public key encryption procedure in $ \Z_p^\ast$}
%--------------------------------------------------------------------
\index{ElGamal!public key}
By varying the Diffie-Hellman key agreement protocol \index{Diffie-Hellman} 
\index{Encryption!ElGamal public key} slightly, you can obtain an 
asymmetric encryption algorithm. This observation was made by Taher ElGamal.
\begin{center}
\fbox{\parbox{12cm}{        
Let $ p $ be a prime number such that the discrete logarithm in $ \Z_p $ is 
difficult to compute. Let $ \alpha \in \Z_p^\ast $ be a primitive element. Let $a \in \N$ and $ 
\beta = \alpha^{a}  \mod p. $\\
\underline{Public key:}
\[ p,\alpha,\beta. \]
\underline{Private key:}
\[a. \]
Let $ k \in \Z_{p-1} $ be a random number and $ x \in \Z_p^{\ast} $ the 
plaintext. \\
\underline{Encryption:}
\[ e_T(x,k)=(y_1,y_2), \]
where
\[ y_1=\alpha^k \mod p\]
and
\[ y_2 = x\beta^k \mod p.\]
\underline{Decryption:}
\[ d_T(y_1,y_2)= y_2 (y_1^{a})^{-1} \mod p \]
}}
\end{center}

@@

%--------------------------------------------------------------------
\subsubsection{Generalised ElGamal public key encryption procedure}
%--------------------------------------------------------------------

>>>>

\newpage
%********************************************************************
\hypertarget{Chapter_ModernCryptography}{}
\section{The Mathematical Ideas behind Modern Cryptography}
\label{Chapter_ModernCryptography}
%********************************************************************
(Oyono R. / Esslinger B., Sep. 2000; Updates Nov. 2000, Feb. 2003)

%--------------------------------------------------------------------
\hypertarget{OneWayFunktion1}{}
\subsection{One way functions with trapdoor and complexity classes}
%--------------------------------------------------------------------
\index{Cryptography!modern} \index{One way function}
A {\bf one way function} is a function that can be calculated 
efficiently, but whose inverse is extremely complicated and practically 
impossible to calculate.\par

@@

\newpage
%--------------------------------------------------------------------
\subsection{Knapsack problem as a basis for public key procedures}%
\index{Cryptography!public key}
%--------------------------------------------------------------------

%--------------------------------------------------------------------
\subsubsection{Knapsack problem}
\index{Knapsack}

@@

%--------------------------------------------------------------------
\subsection{Decomposition into prime factors as a basis for public key 
procedures}\index{Prime factor!decomposition}

%--------------------------------------------------------------------
\hypertarget{RSAVerfahren}{}
\subsubsection[The RSA procedure]
{The RSA procedure\footnotemark}
\footnotetext{
Using CrypTool\index{CrypTool} you can gain practical experience
with the RSA procedure via the menu {\bf Indiv.Procedures \textbackslash{} 
RSA Cryptosystem \textbackslash{} RSA Demonstration}.
}
\index{RSA} \label{rsaverfahren}

As early as 1978, R. \index{Rivest, Ronald} Rivest, \index{Shamir, Adi} 
A. Shamir, \index{Adleman, Leonard} L. Adleman \cite{RSA1978} introduced 
the most important asymmetric cryptography procedure to date.  \par

@@

If we know the various prime factors $ p_1, \dots , p_k $ of $ N, $ then $ \phi 
(N) = N \cdot (1-\frac{1}{p_1}) \,
\cdots \, (1-\frac{1}{p_k}) $\footnote{Further formulas for the Euler phi
function are in the article ``Introduction to Elementary Number Theory with
Examples'', chapter~\ref{patternsandstructures}.}. %or \ref{L-Euler-Function}

@@

At the moment, no algorithm is known that can factorise two prime numbers 
sufficiently quickly for extremely large values (e.g. for several hundred 
decimal places). The quickest algorithms known today
\cite{4Stinson1995}\index{Stinson 1995} 
factorise a compound whole number $ N $ in a time period proportional to $ L(N)= 
e^{\sqrt{\ln (N) \ln (\ln (N))}}. $ 
\vskip +5 pt
\begin{center}
\begin{tabular}{|l||l|l|l|l|l|l|}\hline
$N$ & $ 10^{50} $ & $ 10^{100} $ & $ 10^{150} $ & $ 10^{200} $ & $ 10^{250} $ & 
$ 10^{300} $ \\ \hline
$L(N)$ & $ 1.42 \cdot 10^{10} $ &  $ 2.34  \cdot 10^{15} $ &  $ 3.26 \cdot 
10^{19} $ &  $ 1.20 \cdot 10^{23} $ &  $ 1.86 \cdot 10^{26} $ &  $ 1.53 \cdot 
10^{29} $ \\ \hline
\end{tabular}
\end{center}
\vskip +5 pt

To this date, it has not been proved that the problem of breaking RSA is 
equivalent to the factorisation problem. Nevertheless, it is clear that
the RSA procedure will no longer be safe if the factorisation problem is
``solved''.\footnote{%
In 2000 the authors assumed that values of the order magnitude $ 100 $ to $ 200 $
decimal places are currently safe. They estimates that the current computer
technology indicates that a number with $100$ decimal places could be
factorised in approximately two weeks at justifiable costs, and using an 
expensive configuration (e.g. of around 10 million US dollars), a number with 
$150$ decimal places could be factorised in about a year, and a $200-$digit
number should remain impossible to factorise for a long time to come, unless
there is a mathematical breakthrough. However, you can never be sure that
there won't be a mathematical breakthrough tomorrow.\\
How easy it is to guess the future wrong is shown by the
\hyperlink{RSA-200-chap3}{Factorization of RSA-200} 
(see chapter \ref{NoteFactorisation}) -- completely without a
``mathematical breakthrough''.}

%--------------------------------------------------------------------
\subsubsection{Rabin public key procedure (1979)}

@@

%--------------------------------------------------------------------
\subsection{The discrete logarithm as a basis for public key procedures}
Discrete logarithms \index{Discrete logarithm} form the basis for a large number of algorithms for public-
key procedures.
\index{Logarithm problem!discrete}

%--------------------------------------------------------------------
\subsubsection{The discrete logarithm in $ \Z_p $}
Let $ p $ be a prime number and let $ g \in \Z_p^\ast=\{0,1,\ldots,p-1\} $. Then 
the discrete exponential function base $ g $  is defined as
\[ e_g : k \longrightarrow y:=g^k \mod p, \quad 1\leq k \leq p-1. \]
The inverse function is called a discrete logarithm function $ \log_g $; the 
following holds:
\[ \log_g (g^k) =k. \]
\index{Exponential function!discrete} The problem of the discrete logarithm (in 
$ \Z_p^\ast$) is understood to be as follows:
\[ {\rm Given~ } p,g {\rm ~and~ } y, {\rm~determine~} k {\rm ~such~that~ } 
y=g^k \mod p {\rm .}\]
It is much more difficult to calculate the discrete logarithm than to evaluate 
the discrete exponential function (see chapter \ref{MultOrdPrimitveRoot}).
There are several procedures for calculating the discrete logarithm 
\cite{4Stinson1995}\index{Stinson 1995}:
\vskip + 5pt
\begin{center}
\begin{tabular}{|l|l|}\hline
Name                 &        Complexity \\ \hline \hline
Baby-Step-Giant-Step &         $ O(\sqrt{p}) $ \\ \hline
Silver-Pohlig-Hellman    &    polynomial in $ q, $ the greatest \\
&  prime factor of $ p-1. $ \\ \hline
Index-Calculus &             $ O(e^{(1+o(1)) \sqrt{\ln (p) \ln (\ln (p))}}) $ \\ 
\hline
\end{tabular}
\end{center}
\vskip +5 pt
\index{Silver} \index{Pohlig, S. C.} \index{Hellman, Martin}
\index{Baby-Step-Giant-Step}

@@

%--------------------------------------------------------------------
\subsubsection{ElGamal public key encryption procedure in $ \Z_p^\ast$}
\index{ElGamal!public key}
By varying the Diffie-Hellman key agreement protocol \index{Diffie-Hellman} 
\index{Encryption!ElGamal public key} slightly, you can obtain an 
asymmetric encryption algorithm. This observation was made by Taher ElGamal.
\begin{center}
\fbox{\parbox{12cm}{        
Let $ p $ be a prime number such that the discrete logarithm in $ \Z_p $ is 
difficult to compute. Let $ \alpha \in \Z_p^\ast $ be a primitive element. Let $a \in \N$ and $ 
\beta = \alpha^{a}  \mod p. $\\
\underline{Public key:}
\[ p,\alpha,\beta. \]
\underline{Private key:}
\[a. \]
Let $ k \in \Z_{p-1} $ be a random number and $ x \in \Z_p^{\ast} $ the 
plaintext. \\
\underline{Encryption:}
\[ e_T(x,k)=(y_1,y_2), \]
where
\[ y_1=\alpha^k \mod p\]
and
\[ y_2 = x\beta^k \mod p.\]
\underline{Decryption:}
\[ d_T(y_1,y_2)= y_2 (y_1^{a})^{-1} \mod p \]
}}
\end{center}

@@

%--------------------------------------------------------------------
\subsubsection{Generalised ElGamal public key encryption procedure}
